---
title: "Notes"
format: pdf
editor: visual
toc: true
---re
---

## Module 1 - Jump start

Goals - to understand the packages and workflow

### Importing Files & Examining the data

#### **read_excel**

-   Help to read the excel as tibble

-   Can read specific sheet name or sheet number

-   can handle Na.

-   provide col_types mapping

-   check `?read_excel` for more info.

#### **glimpse**

-   provides overview the data-frame in wide format

**left_join**

-   join the two dataframe base on common variable

### Data Wrangling

separate

-   Split the string column into two or more columns using sep argument

mutate

-   creates new columns that are functions of existing variables.

-   It can also modify (if the name is the same as an existing column) and delete columns (by setting their value to `NULL`)

select

-   Select require columns

-   one provide pattern matching like `endswith` , `startswith` and `contains` or specific datatypes like `is_numeric`

-   `everything()` - list the remaining columns

ggplot2:

-   Refer to ggplot-uncharted course.

## Module 2 - Import & Data wrangling - 1

**KeyPackages - Excel , tidyr, dplyr,**

## Tidyr concept

Long vs Wide:

-   wide: Good for business report, bad for analysis

-   long: Easy for analysis. 99,9% we will tibble format

    Wide ==\> Long

```         
df_tbl |> # gather(key = 'col_name',value = 'value',#list of columns)
```

### Import data

#### CSV :

```         
df <- readr::read_csv(file_path) ## Normal values 
# If there is any problem or warming, one can use 

readr::problems(df)

# Once identified, we can use prase method while reading the file 

df <- readr::read_csv(file_path,col_types = cols(col1 = col_double()))
```

#### RDS :

Save the file as it as file format. it retains the data format.

it can save ggplot files too.

99% for uses cases you can use RDS

```         
df <- readr::read_rds(file_path) 
```

#### EXCEL:

```         
df <- readxl::read_excel(file_path, sheet = "sheet1")
```

### DB Connection

```         
con<- RSQLite::dbConnect(
  drv = SQLite(),
  dbname = 'path_to_databases'
) 


table1_tbl <- tbl(con,'table1_name') |> 
            collect()
            
```

## Data wrangling

### 1. Working with features

-   select

```{r}

# Select simple 
df |> 
select(list_of_col_name_sep_by_comma)

df |> 
select(col1,col2,col3)

# Select helpers 
# starts_with - will select the column that starts with *certain* name 
df |> 
select(starts_with('col_name'))


# starts_with("a"): names that start with "a".
# ends_with("z"): names that end with "z".
# contains("b"): names that contain "b".
# matches("x.y"): names that match regular expression x.y.
# num_range(x, 1:4): names following the pattern, x1, x2, ..., x4.
# all_of(vars)/any_of(vars): matches names stored in the character vector vars # all_of(vars) will error if the variables aren't present; any_of(var) will match just the variables that exist.

# everything(): all variables.

# last_col(): furthest column on the right.

# where(is.numeric): all variables where is.numeric() returns TRUE.

# everything() - remamining column / all variables 


df |> 
    select(col3:col5,everything()) # Rearranging the col3:col5 first and rearranging the remaining
```

## 2. `pull`

**Definition**: Extract a single column as a vector.

```{r}
pull(df, a)   # Returns: 1 2 3
```

------------------------------------------------------------------------

## 3. `distinct`

**Definition**: Return unique rows (optionally by selected columns).

```{r}
df <- tibble(name = c("A", "A", "B"), score = c(1, 1, 2))
distinct(df)                   # Remove duplicate rows
distinct(df, name)            # Unique values of name
```

## 4. `arrange`

**Definition**: Sort rows by column(s).

```{r}

arrange(df, score)            # Ascending
arrange(df, desc(score))      # Descending

```

------------------------------------------------------------------------

## 5. `filter`

**Definition**: Keep rows matching a condition.

```{r}

filter(df, score > 1)
```

------------------------------------------------------------------------

## 6. `slice`

**Definition**: Select rows by position.

```{r}
slice(df, 1)        # First row
slice(df, 1:2)      # First two rows
```

------------------------------------------------------------------------

## 7. `mutate`

**Definition**: Add new columns or modify existing ones

```{r}
mutate(df, score2 = score * 2)
```

------------------------------------------------------------------------

## 8. `summarize` / `summarise`

**Definition**: Reduce data to summary statistics.

```{r}
summarize(df, avg_score = mean(score))
```

------------------------------------------------------------------------

## 9. `group_by` + `summarize`

**Definition**: Grouped summary statistics.

```{r}
df <- tibble(group = c("A", "A", "B"), score = c(1, 2, 3))
df %>% group_by(group) %>%
  summarize(avg = mean(score))
```

------------------------------------------------------------------------

## 10. `set_names`

**Definition**: Set or change column names.

```{r}
df <- tibble(x = 1:3, y = 4:6)
set_names(df, c("a", "b"))
```

------------------------------------------------------------------------

## 11. `rename_with`

**Definition**: Rename columns using a functions

```{r}
rename_with(df, toupper)     # All column names to uppercase

```

------------------------------------------------------------------------

## 12. `spread`

**Definition**: Convert long to wide format (deprecated in favor of `pivot_wider`).

```{r}
df <- tibble(key = c("x", "y"), value = 1:2)
spread(df, key, value)
```

------------------------------------------------------------------------

## 13. `gather`

**Definition**: Convert wide to long format (deprecated in favor of `pivot_longer`).

```{r}
df <- tibble(a = 1:2, b = 3:4)
gather(df, key = "var", value = "val")
```

------------------------------------------------------------------------

## 14. `left_join`

**Definition**: Merge data frames by common columns (like SQL left join).

```{r}
df1 <- tibble(id = 1:2, name = c("A", "B"))
df2 <- tibble(id = 1:2, score = c(10, 20))
left_join(df1, df2, by = "id")


```

------------------------------------------------------------------------

## 15. `bind_cols`

**Definition**: Combine data frames column-wise.

```{r}
df1 <- tibble(x = 1:2)
df2 <- tibble(y = 3:4)
bind_cols(df1, df2)
```

------------------------------------------------------------------------

## 16. `unite`

**Definition**: Combine multiple columns into one.

```{r}
df <- tibble(first = c("John", "Jane"), last = c("Doe", "Smith"))
unite(df, full_name, first, last, sep = " ")
```

------------------------------------------------------------------------

## Module 3 - Import & Data wrangling - 2

```{r}
# ---- lubridate examples ----

# 1. ymd - convert string to Date object
date_obj <- ymd(date_str)
print(date_obj)

# 2. ymd_hms - convert string to POSIXct datetime
datetime_obj <- ymd_hms(datetime_str)
print(datetime_obj)

# 3. year - extract year
print(year(datetime_obj))  # 2025

# 4. month - extract month (numeric, label, abbreviated)
print(month(datetime_obj))                        # 7
print(month(datetime_obj, label = TRUE))          # Jul
print(month(datetime_obj, label = TRUE, abbr = FALSE))  # July

# ---- stringr examples ----

# Sample string
text <- "  The quick Brown fox Jumps over the lazy Dog.  "

# 5. str_detect - check if substring exists
print(str_detect(text, "fox"))     # TRUE

# 6. str_to_upper - convert to upper case
print(str_to_upper(text))

# 7. str_to_lower - convert to lower case
print(str_to_lower(text))

# 8. str_to_title - capitalize each word
print(str_to_title(text))

# 9. str_c - concatenate strings
str1 <- "Hello"
str2 <- "World"
print(str_c(str1, str2, sep = " "))  # "Hello World"

# 10. str_glue - interpolate variables
name <- "Alice"
score <- 95
print(str_glue("Student {name} scored {score}% on the test."))

# 11. tidyr::separate / str_split - split strings

# Using tidyr::separate
df <- data.frame(full_date = "2025-07-24")
df_separated <- separate(df, full_date, into = c("year", "month", "day"), sep = "-")
print(df_separated)

# Using str_split
str <- "apple,banana,cherry"
split_result <- str_split(str, ",")[[1]]
print(split_result)

# 12. str_trim - trim whitespace
print(str_trim(text))  # Removes leading/trailing spaces

# 13. str_replace - replace first match
print(str_replace(text, "the", "a"))  # replaces first "the" (case-sensitive)

# 14. str_replace_all - replace all matches
print(str_replace_all(text, "the", "a"))  # replaces all "the" (case-sensitive)
```

### Forcats

```{r}


library(tidyverse)
library(forcats)


data(mpg)


# 1.0 fct_reorder — Reorder factor levels by another variable (e.g., mean)

# Reorder car classes by average highway mileage
mpg %>%
  mutate(class = fct_reorder(class, hwy)) %>%
  ggplot(aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "fct_reorder: Class reordered by hwy mileage")

# 2.0 fct_reorder2 — Reorder factor levels using both x and y (often used in line charts)

 # Reorder manufacturer by last recorded hwy value in time series (year)
mpg %>%
  group_by(manufacturer, year) %>%
  summarise(mean_hwy = mean(hwy), .groups = "drop") %>%
  mutate(manufacturer = fct_reorder2(manufacturer, year, mean_hwy)) %>%
  ggplot(aes(x = year, y = mean_hwy, color = manufacturer)) +
  geom_line() +
  labs(title = "fct_reorder2: Reordering based on year and mean_hwy")


# 3.0 as_factor — Convert vector to factor (preserves order)


vec <- c("low", "medium", "high")
f <- as_factor(vec)
print(f)  # Levels: low < medium < high



# 4.0 fct_rev - — Reverse factor level order


mpg %>%
  mutate(class = fct_rev(as_factor(class))) %>%
  ggplot(aes(x = class)) +
  geom_bar() +
  coord_flip() +
  labs(title = "fct_rev: Reversed order of class levels")


# 5.0  fct_lump — Lump infrequent levels into "Other"


mpg %>%
  mutate(manufacturer = fct_lump(manufacturer, n = 5)) %>%
  count(manufacturer, sort = TRUE)



# 6.0 fct_relevel() — Change the Order of Factor Levels Manually


# Reorder to put "suv" first, then "compact"
mpg %>%
  mutate(class = fct_relevel(class, "suv", "compact")) %>%
  ggplot(aes(x = class)) +
  geom_bar() +
  labs(title = "fct_relevel: SUV and Compact first") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

\

## Module 4 - Data Visualization

# ggplot2 Geoms Cheat Sheet

```         
library(ggplot2)
data(mpg)
data(diamonds)
```

## geom_point() --- Scatterplot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```

## geom_line() --- Line plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_line()
```

## geom_bar() --- Bar plot (counts)

``` r
ggplot(diamonds, aes(cut)) +
  geom_bar()
```

## geom_col() --- Bar plot (values)

``` r
df <- data.frame(group=c("A","B","C"), value=c(4,7,3))
ggplot(df, aes(group, value)) +
  geom_col()
```

## geom_histogram() --- Histogram

``` r
ggplot(diamonds, aes(price)) +
  geom_histogram(bins = 30)
```

## geom_density() --- Density plot

``` r
ggplot(diamonds, aes(price)) +
  geom_density()
```

## geom_boxplot() --- Boxplot

``` r
ggplot(mpg, aes(class, hwy)) +
  geom_boxplot()
```

## geom_violin() --- Violin plot

``` r
ggplot(mpg, aes(class, hwy)) +
  geom_violin()
```

## geom_area() --- Area plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_area()
```

## geom_tile() --- Tile / Heatmap

``` r
ggplot(diamonds, aes(cut, color, fill = price)) +
  geom_tile()
```

## geom_hex() --- Hexbin plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_hex()
```

## geom_density_2d() --- 2D density plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_density_2d()
```

## geom_smooth() --- Smooth / Regression

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth()
```

## geom_ribbon() --- Ribbon

``` r
df <- data.frame(
  x = 1:10,
  y = 1:10,
  ymin = (1:10) - 2,
  ymax = (1:10) + 2
)

ggplot(df, aes(x, y)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.3) +
  geom_line()
```

## geom_text() --- Text labels

``` r
ggplot(mpg, aes(displ, hwy, label = class)) +
  geom_text()
```

## geom_label() --- Label with box

``` r
ggplot(mpg, aes(displ, hwy, label = class)) +
  geom_label()
```

## geom_segment() --- Line segments

``` r
df <- data.frame(x=1, xend=5, y=1, yend=4)
ggplot(df) +
  geom_segment(aes(x=x, xend=xend, y=y, yend=yend))
```

## geom_abline() --- Reference line

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_abline(slope=1, intercept=0, color="red")
```

## Module 5 - Function programming

### Two types of functions

#### vectorize functions:

Examples: ymd , ceiling_date

first input varaibles would be X that would be vector dtypes

#### Dataframe fuctions

Examples: Select , mutate , etc ,

first input varaibles would be .data or .tbl that would be vector dtypes

### Purrr functions

```{r}
library(purrr)
```

#### 1. map function

Apply a function to each element of a vector or list.

```{r}
nums <- c(1, 2, 3, 4)
# Method 1 
map(nums, sqrt) # 1,4,9,16 
# Method 2 Anonymous function

map(nums, ~ .x^2)

```

#### 2. map\_\* function

These enforce return types.

##### `map_dbl()`

```{r}
map_dbl(nums, sqrt)
```

##### `map_int()`

```{r}
map_int(nums, ~ .x * 2)
```

##### `map_chr()`

```{r}
map_chr(nums, as.character)
```

#### 3. map2() function

Iterate over **two vectors** simultaneously.

```{r}
x <- c(1, 2, 3)
y <- c(10, 20, 30)

map2(x, y, ~ .x + .y)
```

#### 4. pmap function

Iterate over **multiple inputs** stored in a list or data frame.

```{r}
params <- list(
  a = c(1, 2),
  b = c(10, 20),
  c = c(100, 200)
)

pmap(params, ~ ..1 + ..2 + ..3)
```

#### 5. `keep()` and `discard()`

Filter vectors based on a condition.

```{r}
# Keep 

keep(nums, ~ .x > 2)

# Discard 

discard(nums, ~ .x > 2)



```

#### 6. compact()

```{r}
x <- list(1, NULL, 3, NULL, 5)
compact(x)

```

#### 7. possibly()

```{r}
safe_log <- possibly(log, otherwise = NA_real_)

safe_log(10)
safe_log("a")

```

#### 8. safely()

Return both the result and the error (if any).

```{r}

safe_divide <- safely(function(x) 10 / x)

safe_divide(2)
safe_divide(0)

```

#### 9. reduce()

```{r}

# adding 
reduce(nums, `+`)

# multiply 
reduce(nums, `*`)
```

#### 10. map_dbl

```{r}
df <- tibble::tibble(
  x = list(1:3, 4:6, 7:9)
)
df %>%
  dplyr::mutate(
    mean_x = map_dbl(x, mean)
  )
```

## Module 6 - Customer Segmentation

## Module 7 - RMarkdown Primer

## 
