---
title: "Notes"
format: pdf
editor: visual
toc: true
---re
---

## Module 1 - Jump start

Goals - to understand the packages and workflow

### Importing Files & Examining the data

#### **read_excel**

-   Help to read the excel as tibble

-   Can read specific sheet name or sheet number

-   can handle Na.

-   provide col_types mapping

-   check `?read_excel` for more info.

#### **glimpse**

-   provides overview the data-frame in wide format

**left_join**

-   join the two dataframe base on common variable

### Data Wrangling

separate

-   Split the string column into two or more columns using sep argument

mutate

-   creates new columns that are functions of existing variables.

-   It can also modify (if the name is the same as an existing column) and delete columns (by setting their value to `NULL`)

select

-   Select require columns

-   one provide pattern matching like `endswith` , `startswith` and `contains` or specific datatypes like `is_numeric`

-   `everything()` - list the remaining columns

ggplot2:

-   Refer to ggplot-uncharted course.

## Module 2 - Import & Data wrangling - 1

**KeyPackages - Excel , tidyr, dplyr,**

## Tidyr concept

Long vs Wide:

-   wide: Good for business report, bad for analysis

-   long: Easy for analysis. 99,9% we will tibble format

    Wide ==\> Long

```         
df_tbl |> # gather(key = 'col_name',value = 'value',#list of columns)
```

### Import data

#### CSV :

```         
df <- readr::read_csv(file_path) ## Normal values 
# If there is any problem or warming, one can use 

readr::problems(df)

# Once identified, we can use prase method while reading the file 

df <- readr::read_csv(file_path,col_types = cols(col1 = col_double()))
```

#### RDS :

Save the file as it as file format. it retains the data format.

it can save ggplot files too.

99% for uses cases you can use RDS

```         
df <- readr::read_rds(file_path) 
```

#### EXCEL:

```         
df <- readxl::read_excel(file_path, sheet = "sheet1")
```

### DB Connection

```         
con<- RSQLite::dbConnect(
  drv = SQLite(),
  dbname = 'path_to_databases'
) 


table1_tbl <- tbl(con,'table1_name') |> 
            collect()
            
```

## Data wrangling

### 1. Working with features

-   select

```{r}

# Select simple 
df |> 
select(list_of_col_name_sep_by_comma)

df |> 
select(col1,col2,col3)

# Select helpers 
# starts_with - will select the column that starts with *certain* name 
df |> 
select(starts_with('col_name'))


# starts_with("a"): names that start with "a".
# ends_with("z"): names that end with "z".
# contains("b"): names that contain "b".
# matches("x.y"): names that match regular expression x.y.
# num_range(x, 1:4): names following the pattern, x1, x2, ..., x4.
# all_of(vars)/any_of(vars): matches names stored in the character vector vars # all_of(vars) will error if the variables aren't present; any_of(var) will match just the variables that exist.

# everything(): all variables.

# last_col(): furthest column on the right.

# where(is.numeric): all variables where is.numeric() returns TRUE.

# everything() - remamining column / all variables 


df |> 
    select(col3:col5,everything()) # Rearranging the col3:col5 first and rearranging the remaining
```

## 2. `pull`

**Definition**: Extract a single column as a vector.

```{r}
pull(df, a)   # Returns: 1 2 3
```

------------------------------------------------------------------------

## 3. `distinct`

**Definition**: Return unique rows (optionally by selected columns).

```{r}
df <- tibble(name = c("A", "A", "B"), score = c(1, 1, 2))
distinct(df)                   # Remove duplicate rows
distinct(df, name)            # Unique values of name
```

## 4. `arrange`

**Definition**: Sort rows by column(s).

```{r}

arrange(df, score)            # Ascending
arrange(df, desc(score))      # Descending

```

------------------------------------------------------------------------

## 5. `filter`

**Definition**: Keep rows matching a condition.

```{r}

filter(df, score > 1)
```

------------------------------------------------------------------------

## 6. `slice`

**Definition**: Select rows by position.

```{r}
slice(df, 1)        # First row
slice(df, 1:2)      # First two rows
```

------------------------------------------------------------------------

## 7. `mutate`

**Definition**: Add new columns or modify existing ones

```{r}
mutate(df, score2 = score * 2)
```

------------------------------------------------------------------------

## 8. `summarize` / `summarise`

**Definition**: Reduce data to summary statistics.

```{r}
summarize(df, avg_score = mean(score))
```

------------------------------------------------------------------------

## 9. `group_by` + `summarize`

**Definition**: Grouped summary statistics.

```{r}
df <- tibble(group = c("A", "A", "B"), score = c(1, 2, 3))
df %>% group_by(group) %>%
  summarize(avg = mean(score))
```

------------------------------------------------------------------------

## 10. `set_names`

**Definition**: Set or change column names.

```{r}
df <- tibble(x = 1:3, y = 4:6)
set_names(df, c("a", "b"))
```

------------------------------------------------------------------------

## 11. `rename_with`

**Definition**: Rename columns using a functions

```{r}
rename_with(df, toupper)     # All column names to uppercase

```

------------------------------------------------------------------------

## 12. `spread`

**Definition**: Convert long to wide format (deprecated in favor of `pivot_wider`).

```{r}
df <- tibble(key = c("x", "y"), value = 1:2)
spread(df, key, value)
```

------------------------------------------------------------------------

## 13. `gather`

**Definition**: Convert wide to long format (deprecated in favor of `pivot_longer`).

```{r}
df <- tibble(a = 1:2, b = 3:4)
gather(df, key = "var", value = "val")
```

------------------------------------------------------------------------

## 14. `left_join`

**Definition**: Merge data frames by common columns (like SQL left join).

```{r}
df1 <- tibble(id = 1:2, name = c("A", "B"))
df2 <- tibble(id = 1:2, score = c(10, 20))
left_join(df1, df2, by = "id")


```

------------------------------------------------------------------------

## 15. `bind_cols`

**Definition**: Combine data frames column-wise.

```{r}
df1 <- tibble(x = 1:2)
df2 <- tibble(y = 3:4)
bind_cols(df1, df2)
```

------------------------------------------------------------------------

## 16. `unite`

**Definition**: Combine multiple columns into one.

```{r}
df <- tibble(first = c("John", "Jane"), last = c("Doe", "Smith"))
unite(df, full_name, first, last, sep = " ")
```

------------------------------------------------------------------------

## Module 3 - Import & Data wrangling - 2

```{r}
# ---- lubridate examples ----

# 1. ymd - convert string to Date object
date_obj <- ymd(date_str)
print(date_obj)

# 2. ymd_hms - convert string to POSIXct datetime
datetime_obj <- ymd_hms(datetime_str)
print(datetime_obj)

# 3. year - extract year
print(year(datetime_obj))  # 2025

# 4. month - extract month (numeric, label, abbreviated)
print(month(datetime_obj))                        # 7
print(month(datetime_obj, label = TRUE))          # Jul
print(month(datetime_obj, label = TRUE, abbr = FALSE))  # July

# ---- stringr examples ----

# Sample string
text <- "  The quick Brown fox Jumps over the lazy Dog.  "

# 5. str_detect - check if substring exists
print(str_detect(text, "fox"))     # TRUE

# 6. str_to_upper - convert to upper case
print(str_to_upper(text))

# 7. str_to_lower - convert to lower case
print(str_to_lower(text))

# 8. str_to_title - capitalize each word
print(str_to_title(text))

# 9. str_c - concatenate strings
str1 <- "Hello"
str2 <- "World"
print(str_c(str1, str2, sep = " "))  # "Hello World"

# 10. str_glue - interpolate variables
name <- "Alice"
score <- 95
print(str_glue("Student {name} scored {score}% on the test."))

# 11. tidyr::separate / str_split - split strings

# Using tidyr::separate
df <- data.frame(full_date = "2025-07-24")
df_separated <- separate(df, full_date, into = c("year", "month", "day"), sep = "-")
print(df_separated)

# Using str_split
str <- "apple,banana,cherry"
split_result <- str_split(str, ",")[[1]]
print(split_result)

# 12. str_trim - trim whitespace
print(str_trim(text))  # Removes leading/trailing spaces

# 13. str_replace - replace first match
print(str_replace(text, "the", "a"))  # replaces first "the" (case-sensitive)

# 14. str_replace_all - replace all matches
print(str_replace_all(text, "the", "a"))  # replaces all "the" (case-sensitive)
```

### Forcats

```{r}


library(tidyverse)
library(forcats)


data(mpg)


# 1.0 fct_reorder — Reorder factor levels by another variable (e.g., mean)

# Reorder car classes by average highway mileage
mpg %>%
  mutate(class = fct_reorder(class, hwy)) %>%
  ggplot(aes(x = class, y = hwy)) +
  geom_boxplot() +
  coord_flip() +
  labs(title = "fct_reorder: Class reordered by hwy mileage")

# 2.0 fct_reorder2 — Reorder factor levels using both x and y (often used in line charts)

 # Reorder manufacturer by last recorded hwy value in time series (year)
mpg %>%
  group_by(manufacturer, year) %>%
  summarise(mean_hwy = mean(hwy), .groups = "drop") %>%
  mutate(manufacturer = fct_reorder2(manufacturer, year, mean_hwy)) %>%
  ggplot(aes(x = year, y = mean_hwy, color = manufacturer)) +
  geom_line() +
  labs(title = "fct_reorder2: Reordering based on year and mean_hwy")


# 3.0 as_factor — Convert vector to factor (preserves order)


vec <- c("low", "medium", "high")
f <- as_factor(vec)
print(f)  # Levels: low < medium < high



# 4.0 fct_rev - — Reverse factor level order


mpg %>%
  mutate(class = fct_rev(as_factor(class))) %>%
  ggplot(aes(x = class)) +
  geom_bar() +
  coord_flip() +
  labs(title = "fct_rev: Reversed order of class levels")


# 5.0  fct_lump — Lump infrequent levels into "Other"


mpg %>%
  mutate(manufacturer = fct_lump(manufacturer, n = 5)) %>%
  count(manufacturer, sort = TRUE)



# 6.0 fct_relevel() — Change the Order of Factor Levels Manually


# Reorder to put "suv" first, then "compact"
mpg %>%
  mutate(class = fct_relevel(class, "suv", "compact")) %>%
  ggplot(aes(x = class)) +
  geom_bar() +
  labs(title = "fct_relevel: SUV and Compact first") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

\

## Module 4 - Data Visualization

# ggplot2 Geoms Cheat Sheet

```         
library(ggplot2)
data(mpg)
data(diamonds)
```

## geom_point() --- Scatterplot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_point()
```

## geom_line() --- Line plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_line()
```

## geom_bar() --- Bar plot (counts)

``` r
ggplot(diamonds, aes(cut)) +
  geom_bar()
```

## geom_col() --- Bar plot (values)

``` r
df <- data.frame(group=c("A","B","C"), value=c(4,7,3))
ggplot(df, aes(group, value)) +
  geom_col()
```

## geom_histogram() --- Histogram

``` r
ggplot(diamonds, aes(price)) +
  geom_histogram(bins = 30)
```

## geom_density() --- Density plot

``` r
ggplot(diamonds, aes(price)) +
  geom_density()
```

## geom_boxplot() --- Boxplot

``` r
ggplot(mpg, aes(class, hwy)) +
  geom_boxplot()
```

## geom_violin() --- Violin plot

``` r
ggplot(mpg, aes(class, hwy)) +
  geom_violin()
```

## geom_area() --- Area plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_area()
```

## geom_tile() --- Tile / Heatmap

``` r
ggplot(diamonds, aes(cut, color, fill = price)) +
  geom_tile()
```

## geom_hex() --- Hexbin plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_hex()
```

## geom_density_2d() --- 2D density plot

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_density_2d()
```

## geom_smooth() --- Smooth / Regression

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth()
```

## geom_ribbon() --- Ribbon

``` r
df <- data.frame(
  x = 1:10,
  y = 1:10,
  ymin = (1:10) - 2,
  ymax = (1:10) + 2
)

ggplot(df, aes(x, y)) +
  geom_ribbon(aes(ymin = ymin, ymax = ymax), alpha = 0.3) +
  geom_line()
```

## geom_text() --- Text labels

``` r
ggplot(mpg, aes(displ, hwy, label = class)) +
  geom_text()
```

## geom_label() --- Label with box

``` r
ggplot(mpg, aes(displ, hwy, label = class)) +
  geom_label()
```

## geom_segment() --- Line segments

``` r
df <- data.frame(x=1, xend=5, y=1, yend=4)
ggplot(df) +
  geom_segment(aes(x=x, xend=xend, y=y, yend=yend))
```

## geom_abline() --- Reference line

``` r
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_abline(slope=1, intercept=0, color="red")
```

## Module 5 - Function programming

### Two types of functions

#### vectorize functions:

Examples: ymd , ceiling_date

first input varaibles would be X that would be vector dtypes

#### Dataframe fuctions

Examples: Select , mutate , etc ,

first input varaibles would be .data or .tbl that would be vector dtypes

### Purrr functions

```{r}
library(purrr)
```

#### 1. map function

Apply a function to each element of a vector or list.

```{r}
nums <- c(1, 2, 3, 4)
# Method 1 
map(nums, sqrt) # 1,4,9,16 
# Method 2 Anonymous function

map(nums, ~ .x^2)

```

#### 2. map\_\* function

These enforce return types.

##### `map_dbl()`

```{r}
map_dbl(nums, sqrt)
```

##### `map_int()`

```{r}
map_int(nums, ~ .x * 2)
```

##### `map_chr()`

```{r}
map_chr(nums, as.character)
```

#### 3. map2() function

Iterate over **two vectors** simultaneously.

```{r}
x <- c(1, 2, 3)
y <- c(10, 20, 30)

map2(x, y, ~ .x + .y)
```

#### 4. pmap function

Iterate over **multiple inputs** stored in a list or data frame.

```{r}
params <- list(
  a = c(1, 2),
  b = c(10, 20),
  c = c(100, 200)
)

pmap(params, ~ ..1 + ..2 + ..3)
```

#### 5. `keep()` and `discard()`

Filter vectors based on a condition.

```{r}
# Keep 

keep(nums, ~ .x > 2)

# Discard 

discard(nums, ~ .x > 2)



```

#### 6. compact()

```{r}
x <- list(1, NULL, 3, NULL, 5)
compact(x)

```

#### 7. possibly()

```{r}
safe_log <- possibly(log, otherwise = NA_real_)

safe_log(10)
safe_log("a")

```

#### 8. safely()

Return both the result and the error (if any).

```{r}

safe_divide <- safely(function(x) 10 / x)

safe_divide(2)
safe_divide(0)

```

#### 9. reduce()

```{r}

# adding 
reduce(nums, `+`)

# multiply 
reduce(nums, `*`)
```

#### 10. map_dbl

```{r}
df <- tibble::tibble(
  x = list(1:3, 4:6, 7:9)
)
df %>%
  dplyr::mutate(
    mean_x = map_dbl(x, mean)
  )
```

## Module 6 - Customer Segmentation

### K-means

1.  select the number of cluster you want to identify in your data ==\> "K"
2.  Randomly select 3 distinct data points. Initial clusters point
3.  Measure the distance between each point and 3 initial point.
4.  Assign the point to cluster with shortest distance (For 1D it is distance of line , for more than 1 ,it is euclidean distance)
5.  calculate the mean and variance per clusters
6.  repeat steps 2 to 5 , calculate the variations . selected lowest variation
7.  repeat steps 1 to 6 for different value of K. Identify the value of k that provides greater reduction in variation (elbow point)

::: callout-note
k means vs hierarchical clustering

1.  k means requires initial values of K (number of clusters as input). hierarchical clustering just inform us , pairwise, what two things are most similar
:::

### Hierarchical clustering

compare the data points among each other.

-   comparison is done via euclidean distance, Manhattan between them

-   Identify the shortest distance between points and cluster them.

-   We can use multiple ways to compare clusters and point .

#### 1.Start with individual data points

-   Each data point is its **own cluster**.

#### 2. Measure distance between clusters

-   Calculate how similar or different data points are.

-   Common distance measures:

    -   Euclidean distance

    -   Manhattan distance

#### 3. Merge the closest clusters

-   Find the two clusters that are **most similar**.

-   Merge them into a single cluster.

#### 4. Update distances

-   Recalculate distances between the new cluster and all other clusters.

-   This depends on the **linkage method**, such as:

    -   **Single linkage**: closest points

    -   **Complete linkage**: farthest points

    -   **Average linkage**: average distance

    -   **Ward’s method**: minimizes variance within clusters

#### 5. Repeat merging

-   Keep merging the closest clusters.

-   Continue until:

    -   All points are in one cluster, or

    -   You reach the desired number of clusters.

#### 6. Choose the number of clusters

-   Cut the dendrogram at a certain height.

-   Each cut gives you a different number of clusters.

### PCA

is a technique used to **simplify data** by reducing the number of variables while keeping most of the important information.

#### Step 1: Standardize the data

-   Make sure all features have mean = 0 and standard deviation = 1

-   Important because PCA is affected by scale

------------------------------------------------------------------------

#### Step 2: Compute covariance matrix

-   Shows how variables change together

-   Helps understand relationships between features

------------------------------------------------------------------------

#### Step 3: Calculate eigenvalues and eigenvectors

-   **Eigenvectors** → directions (principal components)

-   **Eigenvalues** → amount of variance captured by each component

------------------------------------------------------------------------

#### Step 4: Sort components

-   Sort eigenvalues from largest to smallest

-   The first principal component captures the most variance

------------------------------------------------------------------------

#### Step 5: Select number of components

-   Choose top *k* components that explain most of the variance

-   Often based on:

    -   Explained variance ratio

    -   Scree plot

------------------------------------------------------------------------

#### Step 6: Transform the data

-   Project original data onto selected principal components

-   Get a new reduced dataset

### UMAP

UMAP (Uniform Manifold Approximation and Projection) is a powerful, **non-linear dimensionality reduction algorithm** used to visualize high-dimensional datasets in 2D or 3D, revealing underlying structures like clusters of similar data points.

-   **Visualization:** Its primary use is to plot complex data onto a simple 2D map, showing relationships.

-   **Dimensionality Reduction:** It reduces many variables (dimensions) to just two (UMAP1 & UMAP2) for plotting.

-   **Speed & Scalability:** Generally faster and better at preserving global data structure than similar techniques like t-SNE.

-   **Versatility:** Can be used for general visualization, clustering, or as a pre-processing step for other machine learning tasks. 

### Machine learning

Introduction:

Machine learning is all about to prediction or classifications

Lingo for Machine learning:

Training data - Given data or original data

We use new data (Testing data) to compare the different models (Functions) for identifying the best model for training data.

We do this comparison by defining the cost functions.

-   **Regression Functions (for continuous outputs):**

    -   **Mean Squared Error (MSE):** Averages squared errors, penalizing large errors heavily.

    -   **Mean Absolute Error (MAE):** Averages absolute errors, robust to outliers.

-   **Classification Functions (for categorical outputs):**

    -   **Cross-Entropy:** Measures discrepancy between predicted probabilities and true labels, often using Sigmoid for binary or Softmax for multi-class

::: callout-note
Fitting the Training date well but making poor prediction is called Bias -Variance tradeoff
:::

#### Bias Variance

The inability of machine learning technique to capture the true relationship between independent and dependent is called **Bias**

The difference in fits between training and testing data is called **Variance**.

**Overfit if we have low bias and high variance.**

Simple model will have high bias and low variance

Complex model will have low bias and high varaince.

Three common methods between simple and complex model is **regularization, boosting and bagging.**

#### Cross Validation:

Learning the parameters of a prediction function and testing it on the same data is a methodological mistake: a model that would just repeat the labels of the samples that it has just seen would have a perfect score but would fail to predict anything useful on yet-unseen data. This situation is called **overfitting**. To avoid it, it is common practice when performing a (supervised) machine learning experiment to hold out part of the available data as a **test set.**

There is two different cross validation

-   K-fold cross validation - divide the data into k sample , take k-1 as training data and remaining one sample as testing data. Repeat the above method for k times.

-   Leave One Cross validation: In extreme case, we leave one entire individuals.

#### Linear Regression (simple) :

Def: fitting a line to the data or least square.

y =\> a \* X + b

1.  fit a line that follows random values for a and b.
2.  Identify the Sum of Sqaure Residuals , i.e Total value of squared difference of real value from line
3.  repeated step 1 and 2 till we identify the minimum value for Sum of Sqaure Residuals. Hence this method is called Least Square Method
4.  R^2^ (R squared)

#### Linear Regression (Multiple)

Linear models are amazingly easy to interpret provided we understand how a linear modeling equation works. The prediction function looks like this:

*y_pred = Intercept + c1 x Feature1 + c2 x Feature2 + ... + cN x FeatureN*

Where the:

-   **Intercept:** Is what all predictions start out at.

-   **Model Coefficients (c1 through cN):** Estimates that adjust the features weight in the model

-   **Features (1 through N):** The value of the Feature in the model (categorical are encoded as binary, numeric are left as numeric)

#### Ridge Regression:

when the sample size is small , then ridge regression can improve prediction by making the data less sensitive to the training data.

It is one of the way of regularization.

we introduce small amount of bias while training the models.

instead of minimize the Sum of Sqaure Residuals ,\
in ridge regression , we try to reduce Sum of Sqaure Residuals + lambda \* slope ^\^ 2^

#### **LASSO Regression**

Instead of square the slope in penalty, we take absolute value of slope, then we get Lasso Regression.

Since Lasso regression can exclude useless variables from equations, it is litter better than Ridge regression at reducing the variance in models that contains a lot of useless variables.

#### **Elastic Regression**

For penalty, we combine Lasso Penalty + Ridge Penalty

It does a better job dealing with correlated parameters.

#### Decision tree

decision tree ask a questions and then classify based on the answer given.

very top of the tree is called root node (just the root)

remaining the nodes are called internal node.

very end of the tree is called leaf.

Gini - the technique that uses impurity level and determine which nodes to select and which order (number of yes and number of no) .

Gini for a leaf is calculated = 1- (Probability of Yes) \^ 2 - (Probability of No) \^ 2

Gini for one node = weighted average of gini impurity for leaf nodes.

reepeat for each dimensions (Node)

we select the lowest gini among the nodes.

#### Random Forest

We bootstrap the given data (random sampling) into n times and create decision tree for each bootstrapped data.

Bootstrapping along aggregating many using a technique called "Bagging".

Out of bag samples ==\> are the data that were not present during bootstrapping.

We can measure accurate of random forest by the proportion of out of bag samples that were correctly classified by random forest. (Called as out of bag error)

#### XGBoost:

XGBoost builds many small **decision trees.**

Instead of building one big tree, XGBoost builds trees **one at a time**, and each new tree:

-   Focuses on the **mistakes** made by the previous trees

-   Tries to correct those mistakes

This process is called **gradient boosting**.

Combine weak learners into strong learners ??

##### Gradient Descent

1.  Pick a random values for weights and bias

2.  Then we evaluate how well the models fits the data. i.e. calculate loss function - Sum of square residuals

3.  Repeat the different values for weight bias and calculate of the loss functions

##### AdaBoost: 

Adaboost combines many weak learns to make classifications.

Each weak learns is make by taking the previous stumps' mistakes

#### Coding - Tidy models

1.  parsnip

    contains wrapper for many machine learning models. It provides easy to use API (Similar to scikit learn).

2.  rsample

    Helps to create training testing split.

3.  recipes

    for pre-processing pipelines.

    -   recipe  - Initial setups

    -   step\_ \* - multi steps that can be used for preprocessing

    -   prep

    -   bake

4.  yardstick

    helps us to create metrics for any machine learning algorithms

## Module 7 - RMarkdown Primer

## 
